---
title: "Shortening Psychological Scales Using Computerized Adaptive Testing"
description: |
  A short description of the post.
  ```{r, include=FALSE}
  bytes <- file.size("shortening-psychological-scales-using-computerized-adaptive-testing.Rmd")
  words <- bytes/10
  minutes <- words/200
  ``` 
  (`r round(minutes)` min read)
author:
  - name: Okan Bulut
    url: http://www.okanbulut.com/
    affiliation: University of Alberta
    affiliation_url: https://www.ualberta.ca
    orcid_id: 0000-0001-5853-1267
date: 02-16-2021
categories:
  - psychometrics
  - psychological scales
  - CAT
preview: computer.jpg
bibliography: cat.bib
csl: apa.csl
output:
  distill::distill_article:
    self_contained: false
    toc: true
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
suppressWarnings({
  library("rmarkdown")
  library("kableExtra")
  library("emo")
  library("mirt")
  library("mirtCAT")
  library("DataExplorer")
  library("psych")
  library("ggplot2")
})
```

![Photo by [Glenn Carstens-Peters](https://unsplash.com/@glenncarstenspeters) on [Unsplash](https://unsplash.com/)](computer.jpg)

## Introduction

In my previous posts, I demonstrated how to shorten measurement instruments automatically using methods such as [automated test assembly](https://okan.cloud/posts/2021-01-04-how-to-shorten-a-measurement-instrument-automatically-part-i/) and [the ant colony optimization](https://okan.cloud/posts/2021-01-19-how-to-shorten-a-measurement-instrument-automatically-part-ii/). These methods can help researchers and practitioners build a shorter version of a measurement instrument (e.g., tests, psychological scales, or surveys). 


## Computerized Adaptive Testing

xxx

## Example

In this example, we will use real data from [the Taylor Manifest Anxiety Scale](https://openpsychometrics.org/tests/TMAS/) (TMA) to build a computerized adaptive scale. 


```{r, eval=FALSE}
# Import the TMA data into R
data <- read.csv("tma_data.csv", header = TRUE)

# Preview the data
head(data)
```

```{r, eval=TRUE, echo=FALSE}
data <- read.csv("tma_data.csv", header = TRUE)
paged_table(data, options = list(cols.print = 12))
```

Next, ...

```{r, eval = FALSE}
library("DataExplorer")

plot_correlation(data)
```

```{r, eval = TRUE, echo=FALSE}
plot_correlation(data)
```

Also, ...

```{r, eval = FALSE}
library("psych")
library("ggplot2")

item_summary <- data.frame(items = factor(colnames(data), levels = colnames(data)),
                           difficulty = alpha(data)$item.stats$mean,
                           discrimination = alpha(data)$item.stats$r.cor)

ggplot(data = item_summary, 
       aes(x = items, y = discrimination, colour = discrimination > 0.20)) +
  geom_point(size = 3) + ylim(0, 1) +
  geom_hline(yintercept = 0.20, linetype="dashed", colour = "red", size = 1) +
  labs(x = "Items", "Item-Total Correlation") +
  coord_flip() +
  theme_bw()
```

## Conclusion

xx
