---
title: "How to Shorten a Measurement Instrument Automatically (Part II)"
description: |
  Do you think there are too many questions on your survey? Are you worried that participants 
  may get tired of responding to the questions in the middle of the survey? In this two-part series, I 
  demonstrate how to shorten measurement instruments such as surveys automatically in R. The second part 
  focuses on the use of two optimization algorithms (genetic algorithm and ant colony optimization) for
  reducing the number of questions in surveys and similar instruments.
  ```{r, include=FALSE}
  bytes <- file.size("how-to-shorten-a-measurement-instrument-automatically-part-ii.Rmd")
  words <- bytes/10
  minutes <- words/200
  ``` 
  (`r round(minutes)` min read)
author:
  - name: Okan Bulut
    url: http://www.okanbulut.com/
    affiliation: University of Alberta
    affiliation_url: https://www.ualberta.ca
    orcid_id: 0000-0001-5853-1267
  - name: Hatice Cigdem Bulut
    affiliation: Cukurova University
    affiliation_url: https://www.cu.edu.tr
    orcid_id: 0000-0003-2585-3686
date: 01-21-2021
categories:
  - psychometrics
  - machine learning
  - optimization
bibliography: reducingitems2.bib
preview: pexels-poranimm.jpg
output:
  distill::distill_article:
    self_contained: false
    toc: true
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
suppressWarnings({
library("rmarkdown")
library("kableExtra")
library("emo")
})
```

![Photo by [Poranimm Athithawatthee](https://www.pexels.com/@poranimm-athithawatthee-284222) on [Pexels](https://www.pexels.com/)](pexels-poranimm.jpg)

## Introduction

In the social and behavioral sciences, researchers often use surveys, questionnaires, or scales to collect data from a sample of participants. Such instruments provide an efficient and effective way to collect information about a large group of individuals.

> Surveys are used to collect information from or about people to describe, compare, or
explain their knowledge, feelings, values, and behaviors. [@fink2015conduct]

Sometimes respondents may get tired of answering the questions during the survey taking process---especially if the survey is very long. This is known as *survey taking fatigue*. The presence of survey taking fatigue can affect response quality significantly. When respondents get tired, they may skip questions, provide inaccurate responses due to insufficient effort responding, or even abandon the survey completely. To alleviate this issue, it is important to reduce survey length properly.

In the [first part of this series](https://okan.cloud/posts/2021-01-04-how-to-shorten-a-measurement-instrument-automatically-part-i/), I demonstrated how to use automated test assembly and recursive feature elimination to automatically shorten educational assessments (e.g., multiple-choice exams, tests, and quizzes). In the second part, I will demonstrate how to use the following optimization algorithmsfor creating shorter versions of surveys and similar instruments:

* Genetic algorithm (GA; `r ji("gene")`)
* Ant colony optimization (ACO; `r ji("ant")`)

### Optimization with GA and ACO

In computer science, a genetic algorithm (GA) is essentially a search heuristic that mimics Charles Darwin’s theory of natural evolution. The algorithm reflects the process of natural selection by iteratively selecting and mating high-quality individuals (i.e., solutions) that are more likely to survive and eliminating weak individuals from generating more weak individuals. This process continues until GA finds an optimal or acceptable solution.

> Genetic algorithms are commonly used to generate high-quality solutions to optimization and search problems by relying on biologically inspired operators such as mutation, crossover and selection. [@mitchell1998introduction]

In the context of survey abbreviation, GA can be used as an optimization tool for finding a subset of questions that maximally captures the variance (i.e., $R^2$) in the original data. @yarkoni2010abbreviation proposed the following cost function for scale abbreviation:

$$Cost = Ik + \Sigma^s_i w_i(1-R^2_i)$$
where $I$ represents a user-specified fixed item cost, $k$ represents the number of items to be retained by GA, $s$ is the number of subscales in the measure, $w_i$ are the weights associated with each subscale, and $R^2$ is the amount of variance in the $i^{th}$ subscale explained by its items. If the cost of retaining a particular item is larger than the loss in $R^2$, then the item is dropped from its subscale (i.e., GA returns a shorter subscale). @yarkoni2010abbreviation demonstrated the use of GA in abbreviating lengthy personality scales and thereafter many researchers have used GA to abbreviate psychological scales (e.g., @crone2020developing, @eisenbarth2015using, @gaarticle1)^[See Yarkoni's blog post on the use of GA at https://www.talyarkoni.org/blog/2010/03/31/abbreviating-personality-measures-in-r-a-tutorial/]. 

Like GA, the ant colony optimization (ACO) is also an optimization method. ACO was first inspired by the collective behavior of Argentine ants called *iridomyrmex humilis* [@goss1989self]. While searching for food, ants deposit pheromone on the ground and follow pheromone previously deposited by other ants. Since the shortest path is more likely to retain pheromone, ants are able to follow this path and find promising food sources more quickly. Figure 1 illustrates this process. 

![**Figure 1:** How real ants find a shortest path. (a) Ants arrive at a decision point. (b) Some ants choose the upper path and some the lower path. The choice is random. (c) Since ants move at approximately a constant speed, the ants which choose the lower, shorter, path reach the opposite decision point faster than those which choose the upper, longer, path. (d) Pheromone accumulates at a higher rate on the shorter path. The number of dashed lines is approximately proportional to the amount of pheromone deposited by ants (Obtained from page 54 in @dorigo1996ant)](aco.png)

Engineers decided to use the way Argentine ant colonies function as an analogy to solve the shortest path problem and created the ACO algorithm [@dorigo1996ant; @dorigo1997ant]. Then, @marcoulides2003model applied ACO to model specification searches in structural equation modeling (SEM). The goal of this approach is to automate the model fitting process in SEM by starting with a user-specified model and then fitting alternative models to fix missing paths or parameters. This iterative process continues until an optimal model (e.g., a model with good model-fit indices) is identified. @leite2008item used the ACO algorithm for the development of short forms of scales and found that ACO outperformed traditionally used methods of item selection. In a more recent study, @falk demonstrated how to use ACO for model specification searches in R. 


## Example

In this example, we will use the [Experiences in Close Relationships Scale](http://labs.psychology.illinois.edu/~rcfraley/measures/brennan.html) [@brennan1998self]. The [original data set](http://openpsychometrics.org/_rawdata/ECR-data-1March2018.zip) is available on the Open-Source Psychometric Project [website](https://openpsychometrics.org/). For demonstration purposes, we will use a subset of the original data set based on the following rules: 

* Respondents must participate the survey from the United States, and
* Respondents must be between 18 and 30 years of age. 

In addition, some items have been reverse-coded (items 3, 15, 19, 22, 25, 29, 31, 33, 35) and the participants with missing responses have been eliminated from the data set. The final data set is available [here](https://raw.githubusercontent.com/okanbulut/tds/main/scale_abbreviation/ecr_data.csv). 


### Genetic Algorithm 

We will use the **GAabbreviate** package [@gaabbreviate] to implement GA for scale abbreviation. 

### Ant Colony Optimization

We will use the **ShortForm** package [@shortform] to implement ACO for scale abbreviation. 


## Conclusion

xxx
















